Relationships in the extremes and their influence on competition and coexistence

### 2022-07-30 paper/Paper.Rnw
edits

%Intro to storage effects and why they might be influenced by ATAs
In this paper, we show how ATAs can also be important for competition and for the coexistence of
competing species. In particular, we show how ATAs can strongly influence one specific coexistence mechanism:
storage effects.
%It is straightforward to see, intuitively, why ATAs may influence one mechanism of coexistence in particular: the storage effect. 
%We review, conceptually, what the storage effect is \citep{Chesson_1994,Chesson_2012,Ellner_2016,Barabas_2018}, and explain why ATAs may be relevant to them.
Storage effect allows competing species to fluctuate %over time while escaping extinction because storage effects provide 
in abundance without any going extinct by providing
a relative benefit to whichever species is currently rare; this is the characteristic that makes the storage effect a so-called \emph{stabilizing} mechanism 
of coexistence \citep{Chesson_2000}.
Storage effect can occur if three conditions are met \citep{Chesson_1994}. First, there must be positive 
covariation between the fluctuating quality of the environment (E)
for a species and the strength of competition (C) experienced by that species (called EC covariance), that is stronger when the species is at stable-state abundance than when it is rare. 
Second, growth responses to the environment must be species-specific, so that distinct species do not find the 
same environmental conditions to be optimal. 
Third, a buffering mechanism %must exist whereby species limit 
which limits the influence of competition %in times 
when the environment is poor. 
Dormant stages such as seed banks are an example buffering mechanism \citep{Chesson_2012}. 
When present, these three factors combine so that EC covariance provides a relative benefit to whatever species is currently rare.
Critically, the storage effect can be quantified in invasion analysis by measuring to what extent EC covariance contributes to differences between 
invader (rare) and resident (stable-state) species growth rates \citep{Chesson_1994,Ellner_2016}. %invader and resident 
Thus ATAs may, in principle, influence storage effect simply because storage effect hinges on EC covariation, and ATAs, when they 
occur between environment and competition variables, can be an aspect of EC covariation.

@@

We here pursue goal G1 from the Introduction: to augment MCT to illuminate how ATA
plays a role in coexistence. We develop each step of our new theory for a general 
two-species model (below), and also specifically for the lottery model of the Introduction 
(SI section \ref{SI-SIsect:TheoryForLottery}).
MCT and its recent computational extensions \citep{Ellner_2016,Ellner_2019} 
quantify the contributions of 
multiple mechanisms to coexistence. But we show
that, for purely temporal variation, only one mechanism relates to tail associations: the storage effect. 
So we begin by defining the storage effect. Our mathematical notation parallels
that of \cite{Ellner_2016,Ellner_2019}. It is assumed that the growth rate $r_i(t)=\ln(N_i(t+1)/N_i(t))$ 
of species $i$ can be written as an increasing function of
an environment-dependent factor $E_i(t)$ and as a decreasing function of a quantity representing
competitive pressure on species $i$, $C_i(t)$. We also assume, for simplicity, 
that the $E_i(t)$ are iid through time, and likewise for the
$C_i(t)$. Though the iid assumption %can be circumvented
is not neccesary, %we make the assumption because 
it simplifies the presentation of new ideas.

@@

We now specify how to decompose storage effect into contributions due to ATAs and 
due to $EC$ correlation \emph{per se}. 
Storage effect for species $i$ was computed by comparing
differences between invader and resident species mean growth rates for the original model and for a 
modified model for which 
%the components of $(E_i,C_{i \bs i})$ and those of $(E_j,C_{j \bs i})$
$E$ and $C$
were rendered independent of each other,
for both invader and resideent species [see (\ref{eq:storage_effects_gen})]. 
Because %these components 
$E$ and $C$ 
were rendered completely independent in the modified model, 
storage effect quantifies the contribution of the whole $EC$ relationship, in all its aspects, 
to the invader/resident growth rate difference. 
But we can instead consider distributions $(E_i^\nat,C_{i \bs i}^\nat)$ and 
$(E_j^\nat,C_{j \bs i}^\nat)$ with the same
marginal distributions, respectively, as $(E_i,C_{i \bs i})$ and $(E_j,C_{j \bs i})$,
and with some but not all aspects of the relationship 
between %the components 
$E$ and $C$
altered relative to the original distributions.

Specifically, asymmetries of tail association are removed -- see below for how this is done. 
We then define $\olrn_{i \bs i}$ and $\olrn_{j \bs i}$ based on a modified model 
using  $(E_i^\nat,C_{i \bs i}^\nat)$
and $(E_j^\nat,C_{j \bs i}^\nat)$, and we consider the decomposition 
\begin{align}
\Delta_i^{(EC)} &= [(\olr_{i\bs i} - q_{ij} \olr_{j \bs i})-(\olrn_{i \bs i}-q_{ij} \olrn_{j \bs i})]+
[(\olrn_{i \bs i}-q_{ij} \olrn_{j \bs i})-(\olrs_{i \bs i}-q_{ij} \olrs_{j \bs i})] \label{eq:storage_effects_further1}\\
&= \Delta_i^{[EC]} + \Delta_i^{[E \nat C]}.\label{eq:storage_effects_further}
\end{align}
The term $\Delta_i^{[EC]}$ is the component of storage effect due to 
ATAs between $E_i$ and $C_{i \bs i}$, and between $E_j$ and $C_{j \bs i}$.
The term $\Delta_i^{[E \nat C]}$ is the component of storage effect due to $EC$ 
correlation \emph{per se}, remaining after having removed the effects of ATAs.
The notation $\nat$ was chosen because this new symbol is part of the symbol $\shp$ (i.e., two of its four lines), 
corresponding to the removal of part of the relationship between $E$ and $C$ components (namely, ATAs).
The new symbol should correspondingly be pronounced ``partial sharp''.
The brackets in $\Delta_i^{[E \nat C]}$ and $\Delta_i^{[EC]}$ are intended to indicate that 
(\ref{eq:storage_effects_further}) is a decomposition within a decomposition, i.e., (\ref{eq:storage_effects_further}) is a decomposition
of storage effect, which are themselves a term in the MCT decomposition \citep{Ellner_2019} -- see 
below for additional details.

@@

%multispecies models and multifactor models
\cite{Ellner_2019} elaborated a general theory in which more and different varying factors than
$E$ and $C$ can be considered and used to produce a decomposition of GWR. For instance, if $X$, $Y$ and 
$Z$ are temporally varying factors, \cite{Ellner_2019} produced, for the $i^{\text{th}}$ species, the decomposition
\begin{align}
\text{GWR}_i &= \Delta_i^0+\Delta_i^X+\Delta_i^Y+\Delta_i^Z \label{eq:9}\\
&+\Delta_i^{(X \shp Y)} +\Delta_i^{(XY)} + \Delta_i^{(Y \shp Z)} + \Delta_i^{(YZ)} +\Delta_i^{(X \shp Z)}+\Delta_i^{(XZ)} \label{eq:10}\\
&+ \Delta_i^{(XYZ)}+\Delta_i^{(XY \shp Z)}+\Delta_i^{(X \shp YZ)}+\Delta_i^{(XZ \shp Y)}+\Delta_i^{(X \shp Y \shp Z)}.\label{eq:11}
\end{align}
The terms in (\ref{eq:9}) and (\ref{eq:10}) are analogous to those of (\ref{eq:previous_decomp}).
The new terms in (\ref{eq:11}) result from rendering one or 
more of $X$, $Y$, and $Z$ independent of the others.
We simply comment that this general decomposition should also be extendable to consider the effects of
ATAs in the same way (\ref{eq:full_decomp}) extends (\ref{eq:previous_decomp}), though the complexity
of such a decomposition may present challenges.
For example, $\Delta^{(XY \shp Z)}$ can be decomposed as $\Delta^{[XY \shp Z]} + \Delta^{[X \nat Y \shp Z]}$.
It seems reasonable to suppose that ATAs will play an increasingly important role in systems of 
greater complexity because such systems have more pairs of temporally variable quantities that
may exhibit asymmetries of association. Our prior work shows that ATAs are common in 
ecological and environmental variables \citep{Ghosh_2020_AER}. We have here used simple mathematical
and laboratory models to provide an initial exploration of the potential importance of ATAs
for coexistence. Future work should attempt to generalize lessons learned here to 
field systems with multiple species, stage structure, spatial extent, non-competitive interactions 
such as predation and facilitation that can also mediate coexistence, and other complexities.

>>>>

%Intro to storage effects and why they might be influenced by ATAs
In this paper, we show how ATAs can also be important for competition and for the coexistence of
competing species. In particular, we show how ATAs can strongly influence one specific coexistence mechanism:
storage effects.
%It is straightforward to see, intuitively, why ATAs may influence one mechanism of coexistence in particular: the storage effect. 
%We review, conceptually, what the storage effect is \citep{Chesson_1994,Chesson_2012,Ellner_2016,Barabas_2018}, and explain why ATAs may be relevant to them.
Storage effect allows competing species to fluctuate %over time while escaping extinction because storage effects provide 
in abundance without any going extinct by providing
a relative benefit to whichever species is currently rare; this is the characteristic that makes the storage effect a so-called \emph{stabilizing} mechanism 
of coexistence \citep{Chesson_2000}.
Storage effect can occur if three conditions are met \citep{Chesson_1994}. First, there must be positive 
covariation between the fluctuating quality of the environment (E)
for a species and the strength of competition (C) experienced by that species (called EC covariance), that is stronger when the species is at stable-state abundance than when it is rare. 
Second, growth responses to the environment must be species-specific, so that distinct species do not find the 
same environmental conditions to be optimal. 
Third, a buffering mechanism %must exist whereby species limit 
which limits the influence of competition %in times 
when the environment is poor. 
Dormant stages such as seed banks are an example buffering mechanism \citep{Chesson_2012}. 
When present, these three factors combine so that EC covariance provides a relative benefit to whatever species is currently rare.
Critically, the storage effect can be quantified in invasion analysis by measuring to what extent EC covariance contributes to differences between 
invader (rare) and resident (steady state) species growth rates \citep{Chesson_1994,Ellner_2016}. %invader and resident 
Thus ATAs may, in principle, influence storage effect simply because storage effect hinges on EC covariation, and ATAs, when they 
occur between environment and competition variables, can be an aspect of EC covariation.

@@

We here pursue goal G1 from the Introduction: to augment MCT to illuminate how ATAs
%plays a role in coexistence. 
can structure coexistence. We develop each step of our new theory for a general 
two-species model %(below)
, and also specifically for the lottery model of the Introduction 
(SI section \ref{SI-SIsect:TheoryForLottery}).
MCT and its recent computational extensions \citep{Ellner_2016,Ellner_2019} 
quantify the contributions of 
multiple mechanisms to coexistence. But we show
that, for purely temporal variation, only one mechanism relates to tail associations: the storage effect. 
So we begin by defining the storage effect. Our mathematical notation parallels
that of \cite{Ellner_2016,Ellner_2019}. It is assumed that the growth rate $r_i(t)=\ln(N_i(t+1)/N_i(t))$ 
of species $i$ can be written as an increasing function of
an environment-dependent factor $E_i(t)$ and as a decreasing function of a quantity representing
competitive pressure on species $i$, $C_i(t)$. We also assume, for simplicity, 
that the $E_i(t)$ are iid through time, and likewise for the
$C_i(t)$. Though the iid assumption %can be circumvented
is not necessary, %we make the assumption because 
it simplifies the presentation of new ideas.

@@

We now specify how to decompose storage effect into contributions due to ATAs and 
due to $EC$ correlation \emph{per se}. 
Storage effect for species $i$ was computed by comparing
differences between invader and resident species mean growth rates for the original model and for a 
modified model for which 
%the components of $(E_i,C_{i \bs i})$ and those of $(E_j,C_{j \bs i})$
$E$ and $C$
were rendered independent of each other,
for both invader and resident species [see (\ref{eq:storage_effects_gen})]. 
Because %these components 
$E$ and $C$ 
were rendered completely independent in the modified model, 
storage effect quantifies the contribution of the whole $EC$ relationship, in all its aspects, 
to the invader/resident growth rate difference. 
But we can instead consider distributions $(E_i^\nat,C_{i \bs i}^\nat)$ and 
$(E_j^\nat,C_{j \bs i}^\nat)$ with the same
marginal distributions, respectively, as $(E_i,C_{i \bs i})$ and $(E_j,C_{j \bs i})$,
and with some but not all aspects of the relationship 
between %the components 
$E$ and $C$
altered relative to the original distributions. 
Specifically, asymmetries of tail association are removed. 
We then define $\olrn_{i \bs i}$ and $\olrn_{j \bs i}$ based on a modified model 
using  $(E_i^\nat,C_{i \bs i}^\nat)$
and $(E_j^\nat,C_{j \bs i}^\nat)$, and we consider the decomposition 
\begin{align}
\Delta_i^{(EC)} &= [(\olr_{i\bs i} - q_{ij} \olr_{j \bs i})-(\olrn_{i \bs i}-q_{ij} \olrn_{j \bs i})]+
[(\olrn_{i \bs i}-q_{ij} \olrn_{j \bs i})-(\olrs_{i \bs i}-q_{ij} \olrs_{j \bs i})] \label{eq:storage_effects_further1}\\
&= \Delta_i^{[EC]} + \Delta_i^{[E \nat C]}.\label{eq:storage_effects_further}
\end{align}
The term $\Delta_i^{[EC]}$ is the component of storage effect due to 
ATAs between $E_i$ and $C_{i \bs i}$, and between $E_j$ and $C_{j \bs i}$.
The term $\Delta_i^{[E \nat C]}$ is the component of storage effect due to $EC$ 
correlation \emph{per se}, remaining after having removed the effects of ATAs.
The notation $\nat$ was chosen because this new symbol is part of the symbol $\shp$ (i.e., two of its four lines), 
corresponding to the removal of part of the relationship between $E$ and $C$ components (namely, ATAs).
The new symbol should correspondingly be pronounced ``partial sharp''.
The brackets in $\Delta_i^{[E \nat C]}$ and $\Delta_i^{[EC]}$ are intended to indicate that 
(\ref{eq:storage_effects_further}) is a decomposition within a decomposition, i.e., (\ref{eq:storage_effects_further}) is a decomposition
of storage effect, which are themselves a term in the MCT decomposition \citep{Ellner_2019}.

@@

%multispecies models and multifactor models
\cite{Ellner_2019} elaborated a general theory in which more and different varying factors than
$E$ and $C$ can be considered and used to produce a decomposition of GWR. For instance, if $X$, $Y$ and 
$Z$ are temporally varying factors, \cite{Ellner_2019} produced, for the $i^{\text{th}}$ species, the decomposition
\begin{align}
\text{GWR}_i &= \Delta_i^0+\Delta_i^X+\Delta_i^Y+\Delta_i^Z \label{eq:9}\\
&+\Delta_i^{(X \shp Y)} +\Delta_i^{(XY)} + \Delta_i^{(Y \shp Z)} + \Delta_i^{(YZ)} +\Delta_i^{(X \shp Z)}+\Delta_i^{(XZ)} \label{eq:10}\\
&+ \Delta_i^{(XYZ)}+\Delta_i^{(XY \shp Z)}+\Delta_i^{(X \shp YZ)}+\Delta_i^{(XZ \shp Y)}+\Delta_i^{(X \shp Y \shp Z)}.\label{eq:11}
\end{align}
The terms in (\ref{eq:9}) and (\ref{eq:10}) are analogous to those of (\ref{eq:previous_decomp}).
The new terms in (\ref{eq:11}) result from rendering one or 
more of $X$, $Y$, and $Z$ independent of the others.
We simply comment that this general decomposition should also be extendable to consider the effects of
ATAs in the same way (\ref{eq:full_decomp}) extends (\ref{eq:previous_decomp}), though the complexity
of such a decomposition may present challenges.
For example, $\Delta^{(XY \shp Z)}$ can be decomposed as $\Delta^{[XY \shp Z]} + \Delta^{[X \nat Y \shp Z]}$.

It seems reasonable to suppose that ATAs will play an increasingly important role in systems of 
greater complexity because such systems have more pairs of temporally variable quantities that
may exhibit asymmetries of association. Our prior work shows that ATAs are common in 
ecological and environmental variables \citep{Ghosh_2020_AER}. We have here used simple mathematical
and laboratory models to provide an initial exploration of the potential importance of ATAs
for coexistence. Future work should attempt to generalize lessons learned here to 
field systems with multiple species, stage structure, spatial extent, non-competitive interactions 
such as predation and facilitation that can also mediate coexistence, and other complexities.
